### 2020-08-29 thesis/thesis.tex
[thesis] Spelling

Not only we would like to bring back effects to our type
system~\cite{type-and-effect}, but also do it in a way that is composable. This
is where algebraic effects comes to the rescue. From theoretical point of view,
we need to develop equational theory about our effects to assert correctness of
our language as well as to have the right hammer to reason about our programs.
From practical side, we need to have the way of taming computational effects
in our programs. That is the point where we would like to introduce to unfamiliar
readers a notion of algebraic effects.

@@

        \noindent
        $F_{T}$ is an endofunctor, therefore it sends $X$ into an object in $\mathcal{C}$.
        From any object we have a unique map sending an object into model,
        which is $\eta_{F_{T} (X)}$. From free property of our model we get
        a unique map $\mu{}$ such that the diagram commutes, therefore we have
        a monad $(F_{T}, \eta{}, \mu{})$.

@@

Concluding from the theory that we have built, algebraic effects are just free
models over computation trees, where the signature is the set of effects, and
equational theory is definining rules of rewriting expressions with effects.
The latter is usually irrelevant from implementation perspective, but may be
important for reasoning about programs.

@@

    \begin{lemma}
        In \textbf{Set} category, the following isomorphism holds
        {$\displaystyle \coprod_{b \in B}{A} \cong B \times A $}
    \end{lemma}
    \begin{proof}
        We can think about $\mathVar{B-ary}$ coproducts (tagged unions) over the same set
        such that we have $B$ replicas of $A$ and we select which replica do we want
        to pick. Putting it this way, it immediately follows that it's the same,
        up to isomorphism, as a cartesian product over indexes from $B$ and
        corresponding $A$ sets. For a more detailed and generalized results,
        see~\cite{tensors}.
    \end{proof}

    Model turns into comodel~\cite{comodels}, which we shall also call \textit{world}.
    Putting this together with the above lemma, we obtain the following map:

@@

    We have now derived a dual concept to algebraic effects and handlers, which
    in the literature are also called comodels and runners~\cite{runners-uustalu, runners-in-action}.
    As we have seen, comodels turned into state-passing (co)operations.
    In programming languages, they may be a way to model interaction with external
    resources, where based on current configuration and a parameter, we
    obtain new configuration along with it's result.

@@

    \noindent
    which means that $ \Downarrow{} $ is the \textit{smallest} predicate
    \textit{closed forward} under the above rules. Relation can be inductively
    built from empty set and then in a fixed-point manner we can add more terms.
    Let's now see a relation $ \Uparrow{} \subseteq \Lambda^{0}$ of
    \textit{divergent} terms, which is coinductively defined as follows:

@@

    As can be seen in the J. Yallop repository~\cite{effects-bibliography}, algebraic
    effects and handlers are now trending branch in the programming languages theory.

@@

    From theoretical side, research on comodels and it's relation to state
    dates to papers by John Power et.\ al.~\cite{comodels, tensors}, and on
    the runners of comodels to T. Uustalu~\cite{runners-uustalu}.

@@

It can be seen, that issues lie with the fact, that interacting with coalgebras
may lead to change of their internal state system. Examples include change of
state in NFA, taking next element in an infinite stream or a closing file.
Finalization must occur after initialization, thus we conclude that resumption
should be called at least once. However, modifying state twice the same way can
also lead to unwanted behaviour, as it was seen in
Section~\ref{sec:problem-statement}, therefore it should be invoked exactly
once.

@@

Data-flow analysis may be the right tool for detecting such anomalies, however,
further work in this area is reserved for future.

@@

State passing can be done by embeding given coalgebraic theory in a state monad
over configurations, which in simpler words means, that for a set of coeffects
that we handle, we are going to store current configuration in a state.
Configuration is retrieved before performing a coeffect, and new configuration
is saved after one is handled. Passing of configuration does not leak to the
user, which creates a handy abstraction over operations that manage external
state.

@@

    In the above code, we initialize the coalgebra carrier with filename that
    we want to handle, we open file, write to it once and then close it. Notice
    that file handle is not present in the user code, but only accessible in
    the runner of FileIO as a first element of a pair.

@@

    This example shows how one can implement NFA using coeffects. We abstract
    over internal state of the automata, and based on letters from the alphabet,
    we change automata configuration and return whether word is accepted or not.
    The code below returns whether given NFA accepts word "abba".

@@

    \noindent
    where before executing a coeffect, we obtain current configuration,
    pass it to coeffect, put new config and then return the actual result.
    This is done behind the scenes and user does not know about state
    passing. Note that for readability we have done pattern matching on pairs,
    while currently in Freak it needs to be deconstructed using projections.

@@

    \noindent
    Where the first two are implementing cps for computations and values.
    \verb!cpsOp! is doing translation for an (co)effect that occured in program,
    \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,
    based on a given handler. \verb!forward! is responsible for forwarding the
    computation to the outer handler, and the last one for running the
    translation. This results in an
    implementation that finally supports nested handlers, which can be seen by
    running tests. Links language~\cite{handlers-cps-journal} also supports
    shallow handlers~\cite{shallow-handlers}, whereas our language implements
    only deep ones.

@@

    We have showed derivation of cooperations, based on previous work, as well
    as presented a calculus combining algebraic effects with their dual notion.
    This resulted in a programming language which allows user to seemlessly
    operate on resources based on solid mathematical foundations. It can seen now,
    that (co)algebraic effects are an abstract concept that captures plenty of
    patterns in everyday programming, which already had ad-hoc and more
    restricted implementations in many existing languages, through constructs
    such as \verb!try, except, finally!, or by more low-level mechanisms
    like handling interrupts in operating systems.

@@

    The Links language, on which CPS translation Freak was based, also provides
    small-step operational semantics~\cite{handlers-cps} and an abstract
    machine~\cite{liberating-effects}. Implementing another way of evaluation
    could serve as a way to empirically assert correctness, as opposed to formally.
    Coeffects are living in a syntactic world, which are then translated to
    algebraic effects, thus we may not need to extend the abstract machine to
    accomodate them.

>>>>

Not only we would like to bring back effects to our type
system~\cite{type-and-effect}, but also do it in a way that is composable. This
is where algebraic effects come to the rescue. From theoretical point of view,
we need to develop equational theory about our effects to assert correctness of
our language as well as to have the right hammer to reason about our programs.
From practical side, we need to have the way of taming computational effects
in our programs. That is the point where we would like to introduce to unfamiliar
readers a notion of algebraic effects.

@@

        \noindent
        $F_{T}$ is an endofunctor, therefore it sends $X$ into an object in $\mathcal{C}$.
        From any object we have a unique map sending an object into a model,
        which is $\eta_{F_{T} (X)}$. From free property of our model we get
        a unique map $\mu{}$ such that the diagram commutes, therefore we have
        a monad $(F_{T}, \eta{}, \mu{})$.

@@

Concluding from the theory that we have built, algebraic effects are just free
models over computation trees, where the signature is the set of effects, and
equational theory is defining rules of rewriting expressions with effects.
The latter is usually irrelevant from implementation perspective, but may be
important for reasoning about programs.

@@

    \begin{lemma}
        In \textbf{Set} category, the following isomorphism holds
        {$\displaystyle \coprod_{b \in B}{A} \cong B \times A $}
    \end{lemma}
    \begin{proof}
        We can think about $\mathVar{B-ary}$ coproducts (tagged unions) over the same set
        such that we have $B$ replicas of $A$ and we select which replica we want
        to pick. Putting it this way, it immediately follows that it's the same,
        up to isomorphism, as a cartesian product over indexes from $B$ and
        corresponding $A$ sets. For a more detailed and generalized results,
        see~\cite{tensors}.
    \end{proof}

    Model turns into a comodel~\cite{comodels}, which we shall also call \textit{world}.
    Putting this together with the above lemma, we obtain the following map:

@@

    We have now derived a dual concept to algebraic effects and handlers, which
    in the literature are also called comodels and runners~\cite{runners-uustalu, runners-in-action}.
    As we have seen, comodels turned into state-passing (co)operations.
    In programming languages, they may be a way to model interaction with external
    resources, where based on current configuration and a parameter, we
    obtain a new configuration along with it's result.

@@

    \noindent
    which means that $ \Downarrow{} $ is the \textit{smallest} predicate
    \textit{closed forward} under the above rules. Relation can be inductively
    built from an empty set and then in a fixed-point manner we can add more terms.
    Let's now see a relation $ \Uparrow{} \subseteq \Lambda^{0}$ of
    \textit{divergent} terms, which is coinductively defined as follows:

@@

    As can be seen in the J. Yallop repository~\cite{effects-bibliography}, algebraic
    effects and handlers are now a trending branch in programming languages theory.

@@

    From theoretical side, research on comodels and its relation to state
    dates to papers by John Power et.\ al.~\cite{comodels, tensors}, and on
    the runners of comodels to T. Uustalu~\cite{runners-uustalu}.

@@

It can be seen, that issues lie with the fact, that interacting with coalgebras
may lead to change of their internal state system. Examples include change of
state in NFA, taking the next element in an infinite stream or a closing file.
Finalization must occur after initialization, thus we conclude that resumption
should be called at least once. However, modifying state twice the same way can
also lead to unwanted behaviour, as it was seen in
Section~\ref{sec:problem-statement}, therefore it should be invoked exactly
once.

@@

Data-flow analysis may be the right tool for detecting such anomalies, however,
further work in this area is reserved for the future.

@@

State passing can be done by embedding given coalgebraic theory in a state monad
over configurations, which in simpler words means, that for a set of coeffects
that we handle, we are going to store current configuration in a state.
Configuration is retrieved before performing a coeffect, and new configuration
is saved after one is handled. Passing of configuration does not leak to the
user, which creates a handy abstraction over operations that manage external
state.

@@

    In the above code, we initialize the coalgebra carrier with filename that
    we want to handle, we open a file, write to it once and then close it. Notice
    that file handle is not present in the user code, but only accessible in
    the runner of FileIO as a first element of a pair.

@@

    This example shows how one can implement NFA using coeffects. We abstract
    over internal state of the automata, and based on letters from the alphabet,
    we change automata configuration and return whether word is accepted or not.
    The code below returns whether a given NFA accepts the word "abba".

@@

    \noindent
    where before executing a coeffect, we obtain current configuration,
    pass it to coeffect, put new config and then return the actual result.
    This is done behind the scenes and the user does not know about state
    passing. Note that for readability we have done pattern matching on pairs,
    while currently in Freak it needs to be deconstructed using projections.

@@

    \noindent
    Where the first two are implementing cps for computations and values.
    \verb!cpsOp! is doing translation for an (co)effect that occurred in program,
    \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful continuations,
    based on a given handler. \verb!forward! is responsible for forwarding the
    computation to the outer handler, and the last one for running the
    translation. This results in an
    implementation that finally supports nested handlers, which can be seen by
    running tests. Links language~\cite{handlers-cps-journal} also supports
    shallow handlers~\cite{shallow-handlers}, whereas our language implements
    only deep ones.

@@

    We have showed derivation of cooperations, based on previous work, as well
    as presented a calculus combining algebraic effects with their dual notion.
    This resulted in a programming language which allows user to seamlessly
    operate on resources based on solid mathematical foundations. It can seen now,
    that (co)algebraic effects are an abstract concept that captures plenty of
    patterns in everyday programming, which already had ad-hoc and more
    restricted implementations in many existing languages, through constructs
    such as \verb!try, except, finally!, or by more low-level mechanisms
    like handling interrupts in operating systems.

@@

    The Links language, on which CPS translation Freak was based, also provides
    small-step operational semantics~\cite{handlers-cps} and an abstract
    machine~\cite{liberating-effects}. Implementing another way of evaluation
    could serve as a way to empirically assert correctness, as opposed to formally.
    Coeffects are living in a syntactic world, which are then translated to
    algebraic effects, thus we may not need to extend the abstract machine to
    accommodate them.
